---
title: "Dorothea"
author: "Christian Heyer"
date: "1/15/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE, dev = "png",
  fig.width = 12, fig.height = 12
)
library(dplyr)
library(viper)
library(decoupleR)
library(RColorBrewer)
if (!require(RNAscripts)) {
  devtools::install("./scripts/RNAscripts", upgrade = "never")
}
library(RNAscripts)
library(ggplot2)
library(readr)
library(ComplexHeatmap)
library(DESeq2)
library(purrr)
library(magrittr)
if (exists("snakemake")) {
  ## Input Files
  dds_path <- snakemake@input[["dds_path"]]
  fpkm_path <- snakemake@input[["fpkm_path"]]
  ### Output
  dorothea_fp <- snakemake@input[["dorothea_fp"]]
  # Params
  cond_id <- names(snakemake@config[["diffexp"]][["contrasts"]])
  samp_map <- snakemake@params[["samp_map"]]
  plot_path <- snakemake@params[["plot_path"]]
  comp_groups <- snakemake@config[["signif_comp"]]
  color_scheme <- purrr::map(snakemake@config[["group_colors"]], unlist)
  organism <- snakemake@config[["organism"]]
} else {
  the_yaml <- yaml::read_yaml("../../configs/VascAge_Apelin_config.yaml")

  BASE_ANALYSIS_DIR <- c(the_yaml$dirs$BASE_ANALYSIS_DIR)
  #  comp_id <- "YECplusA-vs-YEC"

  #  contrast_groups <- c("Young-EC_apelin", "Young-EC")
  dds_path <- file.path(paste0(BASE_ANALYSIS_DIR), "deseq2/all.rds")
  fpkm_path <- file.path(BASE_ANALYSIS_DIR, "fpkm/all.tsv")
  samp_map <- the_yaml$samples
  plot_path <- "./"
  cond_id <- names(the_yaml$diffexp$contrasts)
  comp_groups <- the_yaml$comp_groups
  color_scheme <- purrr::map(the_yaml$group_colors, unlist)
  organism <- "Mus musculus"
}
if (!dir.exists(plot_path)) {
  dir.create(plot_path)
}
```

## Dorothea Enrichment

Dorothea is a tool/database containing mouse regulons to infer TF activitiy from gene expression data. Here we use the VIPER method to enrich TF from our RNA-seq data per sample

After loading the data we replace ensembl IDs with gene symbols. We calcuate the TF sites based on the vst transformed expression values from DESeq2. 


```{r read data}
dds_file <- readRDS(dds_path)

Normalized_counts <- getVarianceStabilizedData(dds_file)
# Normalized_counts <- assay(vst(dds_file, blind = F))
fpkm <- read_tsv(fpkm_path)
filer <- fpkm %>%
  dplyr::filter(gene %in% rownames(Normalized_counts)) %>%
  dplyr::filter(!duplicated(gname))

# joined_df <- join_tables(diffexp_tb,filer) %>% dplyr::filter(!duplicated(gname))
Normalized_counts <- Normalized_counts[filer$gene, ]

rownames(Normalized_counts) <- filer$gname

# %>% dplyr::select(gname, stat) %>% dplyr::filter(!is.na(stat)) %>%
# column_to_rownames(var = "gname") %>% as.matrix() -> diffexp_matrix
expression_matrix <- Normalized_counts

# regulons <- dorothea_mm %>% filter(confidence %in% c("A", "B", "C"))
sample_overview <- colData(dds_file) %>% as_tibble(rownames = "sample")
```

## Resulting plots

The first Heatmap shows the Normalized enrichment score for each sample for all transcription factors.
THe condition is given as a colored annotation at the top of the plot. 

```{r pressure, echo=FALSE}
organism_name <- RNAscripts::get_organism_omnipath_name(organism)
net <- get_dorothea(organism = organism_name, levels = c("A", "B", "C"))

# Get the gene names from our expression data
genes_in_data <- rownames(expression_matrix)

# Filter network to only include genes present in our data
net_filtered <- net %>%
  dplyr::filter(target %in% genes_in_data)

cat("Dorothea network filtering:\n")
cat("  Original network:", nrow(net), "interactions\n")
cat("  After filtering:", nrow(net_filtered), "interactions\n")
cat("  Unique TFs:", length(unique(net_filtered$source)), "\n")
cat("  Unique target genes:", length(unique(net_filtered$target)), "\n")

# Check overlap
targets_in_net <- unique(net$target)
overlap <- length(intersect(targets_in_net, genes_in_data))
cat("  Gene overlap:", overlap, "of", length(targets_in_net), 
    "(", round(100*overlap/length(targets_in_net), 1), "%)\n")

if (nrow(net_filtered) == 0) {
  cat("\nWARNING: No overlap between network and data!\n")
  cat("Sample network genes:", head(targets_in_net, 20), "\n")
  cat("Sample data genes:", head(genes_in_data, 20), "\n")
  stop("Network and expression data have no overlapping gene names. Check gene ID types.")
}

net <- net_filtered

tf_activities <- decoupleR::run_viper(expression_matrix, net,
  minsize = 4
)
tf_activities %>%
  dplyr::filter(statistic == "viper") %>%
  tidyr::pivot_wider(
    id_cols = "condition", names_from = "source",
    values_from = "score"
  ) %>%
  tibble::column_to_rownames("condition") %>%
  as.matrix() %>%
  t() -> tf_activities


plot_doro_hm <- function(dds, cond, color_list, p_data) {
  anno_vec <- dds_file@colData %>%
    as.data.frame() %>%
    pull(!!cond)


  color_vec <- color_list[[cond]]
  if (!is.null(color_vec)) {
    ha <- HeatmapAnnotation(
      group = anno_vec,
      col = list(
        group =
          as_vector(color_vec)
      )
    )
  } else {
    ha <- HeatmapAnnotation(group = anno_vec)
  }

  ch <- Heatmap(p_data,
    top_annotation = ha,
    clustering_distance_columns = "euclidean",
    clustering_method_columns = "average",
    show_row_names = FALSE,
    show_column_names = F,
    row_names_gp = gpar(fontsize = 8),
    column_names_rot = 70
  )
  save_cheatmap_svg(ch, file.path(plot_path, glue::glue("progeny_heatmap_{cond}.svg")))

  ch
}

if (!is.null(color_scheme)) {
  purrr::map(cond_id, plot_doro_hm, dds = dds_file, p_data = tf_activities, color_list = color_scheme)
} else {
  purrr::map(cond_id, plot_doro_hm, dds = dds_file, p_data = tf_activities, color_list = color_scheme)
}
```

The following plots show the mean density distribution of the NES for each sample group, followed by a barplot showing the top enrichment for TF activity in each sample group.
The y axis shows the TF factor, the x axis the NES. 

```{r, fig.height = 10}
plot_doro_plots <- function(tf_activities, dds_file, cond) {
  mean_tf_activities <- mean_tibble_from_mat(
    mat = tf_activities,
    contrast_groups = colData(dds_file)[, cond],
    s_map = sample_overview, cond_id = cond
  )
  mean_tf_activities["TF"] <- rownames(tf_activities)
  plot_list <- list()
  for (c_group in unique(sample_overview %>% pull(!!cond))) {
    plot_table <- mean_tf_activities %>%
      dplyr::select(TF, c_group) %>%
      dplyr::arrange(!!sym(c_group))
    histo <- ggplot(plot_table, aes(!!sym(c_group))) +
      geom_density() +
      ggtitle(c_group) +
      theme_bw() +
      xlab("NES")

    plot_table <-
      top_n(
        x = plot_table,
        n = 25,
        wt = abs(!!sym(c_group))
      )
    plot_table <-
      plot_table %>% mutate(color = ifelse(!!sym(c_group) > 0, "green", "red"))

    enrich_plot <-
      ggplot(plot_table, aes(x = TF, y = !!sym(c_group))) +
      scale_x_discrete(limits = plot_table$TF) +
      geom_point(size = 3, aes(col = color)) +
      geom_segment(aes(
        x = TF,
        xend = TF,
        y = 0,
        yend = !!sym(c_group)
      )) +
      coord_flip() +
      theme_bw() +
      ylab("Normalized Enrichment Score (NES)") +
      xlab("TF") +
      ggtitle(c_group)

    NES_plot <-
      ggplot(plot_table, aes(x = reorder(TF, !!sym(c_group)), y = !!sym(c_group))) +
      geom_bar(aes(fill = !!sym(c_group)), stat = "identity") +
      scale_fill_gradient2(
        low = "darkblue",
        high = "indianred",
        mid = "whitesmoke",
        midpoint = 0
      ) +
      theme_minimal() +
      theme(
        axis.title = element_text(face = "bold", size = 12),
        axis.text.x =
          element_text(
            angle = 45,
            hjust = 1,
            size = 10,
            face = "bold"
          ),
        axis.text.y = element_text(size = 10, face = "bold"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none"
      ) +
      ylab("NES") +
      xlab("Pathways") +
      ggtitle(c_group)
    plot(histo)
    plot(enrich_plot)
    plot(NES_plot)
  }
}

purrr::map(cond_id, plot_doro_plots,
  tf_activities = tf_activities,
  dds_file = dds_file
)
```
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## Using decoupleR

```{r}
sample_acts <- run_wmean(
  mat = expression_matrix, network = net, .source = "source", .target = "target",
  .mor = "mor", times = 100, minsize = 5
)

sample_acts_mat <- sample_acts %>%
  dplyr::filter(statistic == "norm_wmean") %>%
  tidyr::pivot_wider(
    id_cols = "condition", names_from = "source",
    values_from = "score"
  ) %>%
  tibble::column_to_rownames("condition") %>%
  as.matrix() %>%
  t()


if (!is.null(color_scheme)) {
  purrr::map(cond_id, plot_doro_hm,
    dds = dds_file,
    p_data = t(scale(t(sample_acts_mat), scale = F)), color_list = color_scheme
  )
} else {
  purrr::map(cond_id, plot_doro_hm, dds = dds_file, p_data = t(scale(t(sample_acts_mat))), color_list = color_scheme)
}
```


```{r, fig.height = 10}
purrr::map(cond_id, plot_doro_plots,
  tf_activities = t(scale(t(sample_acts_mat))),
  dds_file = dds_file
)
```
