---
title: "Visualize Enrichments"
output: html_document
---


# Enrichment analysis of Glmmseq results

Here, the results from Glmmseq GSEA analysis are visualized using the coefficients. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(glmmSeq)
library(ggplot2)
library(RNAscripts)


get_contrast_groups <- function(glmmseq_obj, coef_name) {
    # Split the coefficient name by ":" to check for interaction terms
    coef_terms <- strsplit(coef_name, ":")[[1]]
    
    if (length(coef_terms) > 1) {
        # Interaction term
        interaction_levels <- interaction(glmmseq_obj@metadata[, coef_terms], drop = TRUE, sep = ".")
        coef <- levels(interaction_levels)
        
        # Ensure the reference levels are correctly ordered
        coef <- c(tail(coef, 1), head(coef,1) )
    } else {
        # Single term
        if (is.factor(glmmseq_obj@metadata[, coef_name])) {
            coef <- levels(glmmseq_obj@metadata[, coef_name])
            # Combine non-reference levels and the reference level
            coef <- c(coef[-1], coef[1])
        } else {
            coef <- unique(glmmseq_obj@metadata[, coef_name])
            # Combine non-reference levels and the reference level
        }
    }
    
    coef
}
if (exists("snakemake")) {
    glmmseq_obj <- snakemake@input[["glmmseq_obj"]]
    enrichment_objs <- snakemake@input[["enrichment_obj"]]
    coefs <- snakemake@params[["coefs"]]
    enrichments <- snakemake@params[["enrich_vars"]]
} else {
    glmmseq_obj <-  "/omics/odcf/analysis/OE0228_projects/VascularAging/rna_sequencing/glmmseq/glmmseq/glmmseq_obj.rds.gz"
    enrichment_objs <- c("/omics/odcf/analysis/OE0228_projects/VascularAging/rna_sequencing/glmmseq/glmmseq/cre_status_enrichment_obj.rds.gz",
                         "/omics/odcf/analysis/OE0228_projects/VascularAging/rna_sequencing/glmmseq/glmmseq/EC_status_enrichment_obj.rds.gz",
                         "/omics/odcf/analysis/OE0228_projects/VascularAging/rna_sequencing/glmmseq/glmmseq/age_enrichment_obj.rds.gz",
                         "/omics/odcf/analysis/OE0228_projects/VascularAging/rna_sequencing/glmmseq/glmmseq/age:EC_status_enrichment_obj.rds.gz")
    coefs <- list(
        cre_status = "cre_statusTerf1KO",
        EC_status = "EC_statustumor",
        age = "ageaged",
        "age:EC_status" = "ageaged:EC_statustumor"
    )
    enrichments <- list(
  Hallmark_gsea = list(
    category = "H",
    subcategory = NULL,
    use_gsea = TRUE,
    database = "MSigDB",
    id_class = "ENSEMBL"
  ),
  Reactome_GSEA = list(
    category = "C2",
    subcategory = "CP:REACTOME",
    use_gsea = TRUE,
    database = "MSigDB",
    id_class = "ENTREZID"
  )
)
}

glmmseq_list <- readRDS(glmmseq_obj)
enrichment_objs <- purrr::map(enrichment_objs, readRDS)
names(enrichment_objs) <- names(coefs)


```


## GSEA plots {.tabset}

```{r GSEA_plots, echo=FALSE, fig.width= 12, fig.height = 12, results='asis'}
# Plot GSEA for all enrichments run per object


# Iterate through each coefficient object
for (coef_name in names(coefs)) {
  cat("###", coef_name, "{.tabset}\n\n")
  
  # Get the corresponding enrichment object
  enrichment_obj <- enrichment_objs[[coef_name]]
  
  # Iterate through each GSEA run
  for (enrichment_name in names(enrichments)) {
    cat("####", enrichment_name, "\n\n")
    
    # Get the corresponding GSEA object
    gsea_obj <- enrichment_obj[[enrichment_name]]
    
    # Generate the plot
    plot <- plot_enrichment(gsea_obj$coef@result, X = "ID", Y = "NES", pval = "pvalue",n_max = 20 )
    write(coef_name, file = stderr()) 
    c_groups <- get_contrast_groups(glmmseq_list$norm_counts, coef_name)
    dplots <- RNAscripts::better_dotplot(gsea_obj$coef, c_groups )
    # Print the plot
    print(plot)
    print(dplots)
    cat("\n\n")
  }
}


```


## GSEA Heatmap

Create a Heatmap of all coefficients. Only significant gene sets are shown. 

```{r GSEA Heatmap, fig.width=12, fig.height=12}
library(ComplexHeatmap)
library(circlize)

# Function to retrieve NES scores and adjusted p-values
get_nes_pvalues <- function(gsea_obj) {
    result <- gsea_obj$coef@result
    result$adj_pvalue[result$p.adjust <= 0.05] <- .Machine$double.eps
    return(result)
}

# Define a custom color palette
col_fun <- colorRamp2(c(-2, 0, 2), c("blue", "white", "red"))

# Iterate through each coefficient object
for (enrichment_name in names(enrichments)) {
  data_list <- list()

    for (coef_name in names(coefs)) {
    # Get the corresponding enrichment object
    enrichment_obj <- enrichment_objs[[coef_name]]
    
    # Initialize an empty list to store the data
    
    # Iterate through each GSEA run
        # Get the corresponding GSEA object
        gsea_obj <- enrichment_obj[[enrichment_name]]
        
        # Retrieve NES scores and adjusted p-values
        result <- get_nes_pvalues(gsea_obj)
        
        # Store the data in the list
        data_list[[coef_name]] <- result
    }
    
    # Aggregate the data into a matrix
    gene_sets <- unique(unlist(lapply(data_list, function(x) x$ID)))
    coef_names <- names(data_list)
    nes_matrix <- matrix(0, nrow = length(coef_names), ncol = length(gene_sets), dimnames = list(coef_names, gene_sets))
    
    for (coef_name in coef_names) {
        result <- data_list[[coef_name]]
        nes_matrix[coef_name, result$ID] <- result$NES
    }
    
    # Create the heatmap for the current coefficient
    draw(Heatmap(nes_matrix, name = paste("NES", coef_name), col = col_fun, cluster_rows = TRUE, cluster_columns = TRUE, show_row_names = TRUE, show_column_names = TRUE, row_names_gp = gpar(fontsize = 8),
    column_names_gp = gpar(fontsize = 8)), padding = unit(c(20, 2, 2, 2), "mm"))
}
```



## DecoupleR Based Enrichment results


### Progeny

```{r}

library(decoupleR)
library(tidyr)
library(pheatmap)

# Get progeny and CollectTRI from the decoupleR package
progeny <- decoupleR::get_progeny(organism = "mouse")
progeny <- progeny %>% dplyr::filter(!is.na(target))
collectTRI <- decoupleR::get_collectri(organism = "mouse")

mat <- glmmseq_list$norm_counts@stats$coef[, unlist(coefs)] 
#mat <- mat[,-1]

pval_mat <-  glmmseq_list$norm_counts@stats$pvals[,unlist(names(coefs))]
pval_mat[pval_mat == 0]<- 1e-300
# Run DecoupleR
sample_acts <- run_mlm(mat=mat, net=progeny, .source='source', 
                       .target='target',
                  .mor='weight', minsize = 5)
sample_acts_mat <- sample_acts %>%
  dplyr::mutate(padj = p.adjust(p_value, method = "BH")) %>%
  dplyr::filter(padj <= 0.05) %>%
  pivot_wider(id_cols = 'condition', names_from = 'source',
              values_from = 'score') %>%
  tibble::column_to_rownames('condition') %>%
  as.matrix()
sample_acts_mat[is.na(sample_acts_mat)] <- 0
# Choose color palette
palette_length = 100
my_color = colorRampPalette(c("Darkblue", "white","red"))(palette_length)

my_breaks <- c(seq(-3, 0, length.out=ceiling(palette_length/2) + 1),
               seq(0.05, 3, length.out=floor(palette_length/2)))

# Plot
ComplexHeatmap::Heatmap(sample_acts_mat)
```

### Per Condition plots for pathways

```{r, dpi=300, fig.width=8, fig.height=8}
plot_pathway <- function(net, pathway, deg, cond) {
  df <- net %>%
    dplyr::filter(source == pathway) %>%
    dplyr::arrange(target) %>%
    dplyr::mutate(ID = target, color = "3") %>%
    tibble::column_to_rownames('target')
  
  inter <- sort(dplyr::intersect(rownames(deg), rownames(df)))
  
  df <- df[inter, ]
  
  df['lfc'] <- deg[inter, ]
  
  df <- df %>%
    dplyr::mutate(color = dplyr::if_else(weight > 0 &
                                           lfc > 0, '1', color)) %>%
    dplyr::mutate(color = dplyr::if_else(weight > 0 &
                                           lfc < 0, '2', color)) %>%
    dplyr::mutate(color = dplyr::if_else(weight < 0 &
                                           lfc > 0, '2', color)) %>%
    dplyr::mutate(color = dplyr::if_else(weight < 0 &
                                           lfc < 0, '1', color))
  
  colors <- rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu")[c(2, 10)])
  
  p <- ggplot2::ggplot(data = df,
                       mapping = ggplot2::aes(x = weight, y = lfc, color = color)) +
    ggplot2::geom_point(size = 2.5, color = "black") +
    ggplot2::geom_point(size = 1.5) +
    ggplot2::scale_colour_manual(values = c(colors[2], colors[1], "grey")) +
    ggrepel::geom_label_repel(mapping = ggplot2::aes(label = ID),label.size = 1) +
    ggplot2::theme_minimal() +
    ggplot2::theme(
      legend.position = "none",
      axis.text = ggplot2::element_text(size = 12),       # Axis tick labels
      axis.title = ggplot2::element_text(size = 14)       # Axis titles
    ) +
    ggplot2::geom_vline(xintercept = 0, linetype = 'dotted') +
    ggplot2::geom_hline(yintercept = 0, linetype = 'dotted') +
    ggplot2::ggtitle(pathway)
  
  p
}
# Assuming 'conditions' is a vector of your conditions
conditions <- unique(sample_acts$condition)
sample_acts <- sample_acts %>% dplyr::mutate(padj = p.adjust(p_value, method = "BH"))
# Plot
colors <- rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu")[c(2, 10)])

for (condition in conditions) {
  # Filter data for the current condition
  condition_data <- sample_acts %>%
     dplyr::filter(condition == !!condition) %>% dplyr::filter(padj <= 0.05)
                                      
  #condition_data <- 
  # Plot
  p <- ggplot(condition_data, aes(x = reorder(source, score), y = score)) + 
    geom_bar(aes(fill = score), stat = "identity") +
       ggplot2::scale_fill_gradient2(low = colors[1], 
                                   mid = "whitesmoke", 
                                   high = colors[2], 
                                   midpoint = 0) + 
     ggplot2::theme_minimal() +
     ggplot2::theme(axis.title = element_text(face = "bold", size = 15),
              axis.text.x = ggplot2::element_text(angle = 45, 
                                                  hjust = 1, 
                                                  size = 12, 
                                                  face = "bold"),
              axis.text.y = ggplot2::element_text(size = 12, 
                                                  face = "bold"),
                legend.title = element_text(size = 15),
                legend.text  = element_text(size = 13),
              panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank())+
    xlab("Pathways") +
    ggtitle(condition)  # Add condition as title
  
  print(p)
  single_plots<- purrr::map(condition_data$source, plot_pathway, net = progeny, deg = mat[,condition, drop = F] , cond = condition)
  purrr::map(single_plots, print)
  
}


```



## Enrich TFs with collectTRI


## Enrich against collectTRI
```{r}


# Run ulm
sample_acts <- run_ulm(mat=mat[,], net=collectTRI, .source='source', .target='target',
                  .mor='mor', minsize = 5)

n_tfs <- 25

# Transform to wide matrix
sample_acts_mat <- sample_acts %>% dplyr::mutate(padj = p.adjust(p_value, method = "BH")) %>%
  dplyr::filter(padj <= 0.05) %>%
  pivot_wider(id_cols = 'condition', names_from = 'source',
              values_from = 'score') %>%
  tibble::column_to_rownames('condition') %>%
  as.matrix()

# Get top tfs with more variable means across clusters
tfs <- sample_acts %>% dplyr::mutate(padj = p.adjust(p_value, method = "BH")) %>%
    dplyr::filter(padj <= 0.05) %>%
  dplyr::group_by(source) %>%
  dplyr::summarise(std = sd(score)) %>%
  dplyr::arrange(-abs(std)) %>%
  head(n_tfs) %>%
  dplyr::pull(source)
sample_acts_mat <- sample_acts_mat[,tfs]

# Scale per sample
#sample_acts_mat <- scale(sample_acts_mat)
sample_acts_mat[is.na(sample_acts_mat)] <- 0
# Choose color palette
palette_length = 100
my_color = colorRampPalette(c("Darkblue", "white","red"))(palette_length)

my_breaks <- c(seq(min(sample_acts_mat), 0, length.out=ceiling(palette_length/2) + 1),
               seq(0.05, max(sample_acts_mat), length.out=floor(palette_length/2)))

# Plot

pheatmap(sample_acts_mat, border_color = NA, color=my_color , breaks = my_breaks) 
```

```{r}
# Assuming 'conditions' is a vector of your conditions
conditions <- unique(sample_acts$condition)
sample_acts <- sample_acts %>% dplyr::mutate(padj = p.adjust(p_value, method = "BH"))
for (condition in conditions) {
  # Filter data for the current condition
  condition_data <- sample_acts[sample_acts$condition == condition, ] %>% 
    dplyr::filter(padj < 0.05)
  
  # Filter top TFs in both signs
  f_contrast_acts <- condition_data %>%
    dplyr::mutate(rnk = NA)
  msk <- f_contrast_acts$score > 0
  f_contrast_acts[msk, 'rnk'] <- rank(-f_contrast_acts[msk, 'score'])
  f_contrast_acts[!msk, 'rnk'] <- rank(-abs(f_contrast_acts[!msk, 'score']))
  tfs <- f_contrast_acts %>%
    dplyr::arrange(rnk) %>%
    head(n_tfs) %>%
    dplyr::pull(source)
  f_contrast_acts <- f_contrast_acts %>%
    dplyr::filter(source %in% tfs)

  # Plot
  p <- ggplot(f_contrast_acts, aes(x = reorder(source, score), y = score)) + 
      geom_bar(aes(fill = score), stat = "identity") +
            ggplot2::scale_fill_gradient2(low = colors[1], 
                                   mid = "whitesmoke", 
                                   high = colors[2], 
                                   midpoint = 0) + 
     ggplot2::theme_minimal() +
     ggplot2::theme(axis.title = element_text(face = "bold", size = 15),
              axis.text.x = ggplot2::element_text(angle = 45, 
                                                  hjust = 1, 
                                                  size = 13, 
                                                  face = "bold"),
              axis.text.y = ggplot2::element_text(size = 13, 
                                                  face = "bold"),
                legend.title = element_text(size = 15),
                legend.text  = element_text(size = 13),
              panel.grid.major = element_blank(), 
              panel.grid.minor = element_blank())+
      xlab("TFs") +
      ggtitle(condition)  # Add condition as title
  
  print(p)
}
```

